A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data

### 2014-04-16 bioinformatics/document.tex
Added supplement and updated for alexey comments

\title[Hadoop for analyzing DNA sequence data]{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
\author[Alexey Siretskiy]{Corresponding Author\,$^1$\footnote{to whom correspondence should be addressed}, Luca Pireddu\,$^{2}$ and Ola Spjuth\,$^{3,4}$}
\address{$^{1}$Department of Information Technology, Uppsala University, Uppsala, Sweden\\
$^{2}$CRS4, Polaris, Pula, Italy\\
$^{3}$Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden\\
$^{4}$Science for Life Laboratory, Uppsala University, Uppsala, Sweden}

@@

 \begin{subfigure}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figures/fig1-sep.png}
	\caption{}
		\label{fig:fig1}
        \end{subfigure}%
 \begin{subfigure}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figures/fig2-sep.png}
	\caption{}
	\label{fig:fig2}
        \end{subfigure}
	\caption{
	\textbf{(a)} Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
	increasing input dataset size. Each test dataset (labelled with Roman
	numerals) was used. The image clearly highlights the linear scaling
	characteristics of the Hadoop-based approach.\\
	\textbf{(b)} The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
	a 16-core HPC node, respectively.
	The points are fit to a quadratic least-squares curve, which makes it possible
	to predict {\it infinite} dataset size. Datasets are labeled with Roman
	numerals.
		}
\end{figure*}

@@

\subsection*{Datasets}
For our experiments we used three publicly available DNA-seq datasets~(I--III),
a synthetic dataset~(IV) of {\it A.thaliana} -- the well-known model plant --
and a dataset (V) of two {\it H.sapiens} individuals
(Table~\ref{table:datasets}). Datasets I-III and V are composed of Illumina
HiSeq sequencing data. Additional information about the source of the datasets is
provided in the Supplementary Material section.

@@

\subsection*{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
which is heavily used for analysis of NGS-data \citep{lampa}.
Data and reference genomes were stored on a parallel, shared storage system.
On the other hand, the Hadoop test platform was deployed on a private cloud at
UPPMAX using the OpenNebula (http://opennebula.org) cloud manager. The cluster was
configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0 (http://www.cloudera.com/content/\\cloudera/en/why-cloudera/hadoop-and-big-data.html).
For the fully detailed hardware configuration of the two computational resources
please see the Supplementary Material section.

@@

To quantify the processing time saved by adopting the MR approach for
preprocessing the read data, we ran an
experiment comparing the new preprocessor described in
Section~{\it MapReduce-based Data Preprocessor} to the original preprocessor furnished with
Crossbow. To further validate our strategy, we also compared the performance of
our MapReduce approach to a different parallelization strategy based on a custom Bash script
that uses all processing cores available on a single computing node to accelerate the operation (see Supplementary Material for details of the implementation). This program read its
input directly from conventional HPC storage, rather than HDFS\@.  The experiment
was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
with 56 cores (7 slave nodes) while the Bash approach ran on a single node
equipped with 16 CPU cores.

@@

Since Hadoop was designed to digest huge datasets\citep{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach. In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
$F=T_{p}\times p,$ where~$T_{p}$ is the calculation time on~$p$ cores.

Using the data from Table~\ref{table:pipleline-timings}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}). Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset. We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases. 
The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on 'bare metal' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.

@@

Hadoop presents a computing paradigm that is different from what most users are
accustomed to using.  Therefore, adopting it requires an investment in learning
how it works, how to use it, and how to understand what the programs running on
the platform are doing.
A popular way to define and run bioinformatics pipelines on HPC and cloud resources
is via the Galaxy\citep{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
user interface~(GUI) to bioinformatic programs, simplifying the experience for
the end user.
A similar project is Cloudgene\citep{cloudgene}, which provides a GUI to help lower the learning curve
for Hadoop adopters and is specifically targeted at users in the life
sciences. It consists of a light-weight and flexible Web-based solution for both
public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
extended the platform with functions to import data from the central file system
at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
particular task related to DNA sequencing, Cloudgene makes it easy to select
data and parameters for its execution. Most of the data management work is done
automatically and the results can be downloaded to the client machine. The
modular structure allows modification of the source code to adapt to the
existing computing center architecture. For example, UPPMAX users can import
their data from the sequencing platform directly to the Hadoop cluster by
pressing a button and entering the credentials; at the same time they can be
sure that their sensitive data is kept private.

@@

\section*{Acknowledgement}
We thank system experts Pontus Freyhult, Tore Sundqvist, and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage. We also thank Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.

\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
This work was supported by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) [p2013023]; the Swedish strategic research programme eSSENCE; and COST Action BM1006 Next Generation Sequencing Data Analysis Network SeqAhead. Part of L.P.'s activity was performed within the context of the Ph.D.\ program in Biomedical Engineering at the University of Cagliari, Italy.\\

>>>>

\title[Hadoop for analyzing DNA sequence data]{A quantitative assessment of the Hadoop framework for analyzing massively parallel DNA sequencing data}
\author[Alexey Siretskiy]{Alexey Siretskiy\,$^1$\footnote{to whom correspondence should be addressed}, Luca Pireddu\,$^{2}$, Tore Sundqvist\,$^1$ and Ola Spjuth\,$^{3,4}$}
\address{$^{1}$Department of Information Technology, Uppsala University, Uppsala, Sweden\\
$^{2}$CRS4, Polaris, Pula, Italy\\
$^{3}$Department of Pharmaceutical Biosciences, Uppsala University, Uppsala, Sweden\\
$^{4}$Science for Life Laboratory, Uppsala University, Uppsala, Sweden}

@@

 \begin{subfigure}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figures/fig1-sep.png}
	\caption{}
		\label{fig:fig1}
        \end{subfigure}%
 \begin{subfigure}[b]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figures/fig2-sep.png}
	\caption{}
	\label{fig:fig2}
        \end{subfigure}
	\caption{
	\textbf{(a)} Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
	increasing input dataset size. Each test dataset (labelled with Roman
	numerals) was used. The image clearly highlights the linear scaling
	characteristics of the Hadoop-based approach.\\
	\textbf{(b)} The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
	a 16-core HPC node, respectively.
	The points are fit to a quadratic least-squares curve, which makes it possible
	to predict {\it infinite} dataset size. Datasets are labeled with Roman
	numerals; f.r. stands for "forward reads".
		}
\end{figure*}

@@

\subsection*{Datasets}
For our experiments we used three publicly available DNA-seq datasets~(I--III),
a synthetic dataset~(IV) of {\it A.thaliana} -- the well-known model plant --
and a dataset (V) of two {\it H.sapiens} individuals
(Table~\ref{table:datasets}). Datasets I-III and V are composed of Illumina
HiSeq sequencing data. Additional information about the source of the datasets is
provided in the Supplementary Data.

@@

\subsection*{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
which is heavily used for analysis of NGS-data \citep{lampa}.
Data and reference genomes were stored on a parallel, shared storage system.
On the other hand, the Hadoop test platform was deployed on a private cloud at
UPPMAX using the OpenNebula (http://opennebula.org) cloud manager. The cluster was
configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0 (http://www.cloudera.com/content/\\cloudera/en/why-cloudera/hadoop-and-big-data.html).
For the fully detailed hardware configuration of the two computational resources
please see the Supplementary Data.

@@

To quantify the processing time saved by adopting the MR approach for
preprocessing the read data, we ran an
experiment comparing the new preprocessor described in
Section~{\it MapReduce-based Data Preprocessor} to the original preprocessor furnished with
Crossbow. To further validate our strategy, we also compared the performance of
our MapReduce approach to a different parallelization strategy based on a custom Bash script
that uses all processing cores available on a single computing node to accelerate the operation (see Supplementary Data for details of the implementation). This program read its
input directly from conventional HPC storage, rather than HDFS\@.  The experiment
was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
with 56 cores (7 slave nodes) while the Bash approach ran on a single node
equipped with 16 CPU cores.

@@

Since Hadoop was designed to digest huge datasets \citep{hadoop,lin2010} we hypothesize that the larger the biological sequence dataset is, the more suitable Hadoop becomes compared to the HPC approach. In order to compare the 'suitability' for the Hadoop and HPC platforms run on different number of cores, we construct the function showing the amount of used core-hours:
$F=T_{p}\times p,$ where~$T_{p}$ is the calculation time on~$p$ cores.

Using the data from Table~\ref{table:pipleline-timings}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}). Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset. We note that the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases. 
The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on "bare metal" of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.

@@

Hadoop presents a computing paradigm that is different from what most users are
accustomed to using.  Therefore, adopting it requires an investment in learning
how it works, how to use it, and how to understand what the programs running on
the platform are doing.
A popular way to define and run bioinformatics pipelines on HPC and cloud resources
is via the Galaxy\citep{galaxy,Afgan:2010uq} platform, which provides a Web-based graphical
user interface~(GUI) to bioinformatic programs, simplifying the experience for
the end user.
A similar project is Cloudgene \citep{cloudgene}, which provides a GUI to help lower the learning curve
for Hadoop adopters and is specifically targeted at users in the life
sciences. It consists of a light-weight and flexible Web-based solution for both
public and private clouds. We implemented our Hadoop-pipeline in Cloudgene and
extended the platform with functions to import data from the central file system
at UPPMAX into HDFS on the private cloud (Figure~\ref{fig:fig4}).  For our
particular task related to DNA sequencing, Cloudgene makes it easy to select
data and parameters for its execution. Most of the data management work is done
automatically and the results can be downloaded to the client machine. The
modular structure allows modification of the source code to adapt to the
existing computing center architecture. For example, UPPMAX users can import
their data from the sequencing platform directly to the Hadoop cluster by
pressing a button and entering the credentials; at the same time they can be
sure that their sensitive data is kept private.

@@

\section*{Acknowledgement}
We thank system experts Pontus Freyhult, and Peter Ankerst{\aa}l at UPPMAX for valuable discussions on effective storage and network usage. We also thank Jonas Hagberg (BILS, Stockholm, Sweden) for implementing the Cloudgene extensions to import data from UPPMAX filesystem, as well as Wesley Schaal for proofreading.

\paragraph{Funding\textcolon} This work was supported by SNIC through Uppsala Multidisciplinary Center for Advanced Computational Science (SNIC-UPPMAX) [p2013023]; the Swedish strategic research programme eSSENCE; and COST Action BM1006 Next Generation Sequencing Data Analysis Network SeqAhead. Part of L.P.'s activity was performed within the context of the Ph.D.\ program in Biomedical Engineering at the University of Cagliari, Italy.\\
